{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming\\[0.25em]

### 2025-07-10 paper.tex
Add teaser, edit page 1

% TODO: Figure out how to call those "programming experiences" (also is this a good name?)
%   the text uses "concrete, collaborative, interactive" - but this is a silly list
%   really, they are "cool" but how to say that.
%
% TODO: Also, what exactly goes into the list?
%
%
\documentclass[sigconf,anonymous,screen]{acmart}
%\documentclass[sigconf,review,anonymous]{acmart}

@@

\newcommand{\ident}[1]{{\sffamily #1}}
\newcommand{\srcid}[1]{\textnormal{\sffamily #1}}
\newcommand{\srckvd}[1]{\textnormal{\sffamily\bfseries #1}}
\newcommand{\note}[1]{\textcolor{red}{#1}}

@@

\setcopyright{none}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXX.XXXX}

\acmConference[Conference acronym 'XX]{}{June 03--05, 2018}{Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/18/06}

\begin{document}
\title[Denicek: Computational Substrate for Document-Oriented End-User
  Programming]{{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming}

@@

\begin{abstract}
User-centric programming research gave rise to a variety of compelling programming experiences,
including collaborative source code editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming languages and system.

@@

%\begin{teaserfigure}
%  \includegraphics[width=\textwidth]{sampleteaser}
%  \caption{Seattle Mariners at Spring Training, 2010.}
%  \Description{Enjoying the baseball game from the third-base
%  seats. Ichiro Suzuki preparing to bat.}
%  \label{fig:teaser}
%\end{teaserfigure}
\maketitle

\section{Introduction}

A computational substrate defines the structures with which programs are constructed, how
the program state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
The choice of a substrate affects what programming experiences can be readily supported.
For example, object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory \cite{steele-1993-lisp}.

@@

In this paper, we describe a programming substrate that makes it easy to develop diverse
programming experiences \cite{myers-2006-eup}. The experiences we focus on include:

>>>>

% TODO: Figure out how to call those "programming experiences" (also is this a good name?)
%   the text uses "concrete, collaborative, interactive" - but this is a silly list
%   really, they are "cool" but how to say that.
%
% TODO: Also, what exactly goes into the list?
%
%
\documentclass[sigconf]{acmart}
%\documentclass[sigconf,review,anonymous]{acmart}

@@

\setlength{\marginparwidth}{1.4cm}

\newcommand{\diff}[1]{{\textcolor{magenta}{#1}}}
\newcommand{\note}[1]{\marginpar{\raggedright\scriptsize\textcolor{magenta}{#1}}}

\newcommand{\ident}[1]{{\sffamily #1}}
\newcommand{\srcid}[1]{\textnormal{\sffamily #1}}
\newcommand{\srckvd}[1]{\textnormal{\sffamily\bfseries #1}}

@@

\copyrightyear{2025}
\acmYear{2025}
\setcopyright{cc}
\setcctype{by}
\acmConference[UIST '25]{The 38th Annual ACM Symposium on User Interface Software and Technology}{September 28-October 1, 2025}{Busan, Republic of Korea}
\acmBooktitle{The 38th Annual ACM Symposium on User Interface Software and Technology (UIST '25), September 28-October 1, 2025, Busan, Republic of Korea}\acmDOI{10.1145/3746059.3747646}
\acmISBN{979-8-4007-2037-6/2025/09}

\begin{document}
\title[Denicek: Computational Substrate for Document-Oriented End-User
  Programming]{{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming\\[0.25em]}

@@

\author{Jonathan Edwards}
\email{jonathanmedwards@gmail.com}
\orcid{0000-0003-1958-7967}
\affiliation{%
  \institution{Independent}
  \city{Boston}
  \country{USA}
}

\begin{abstract}
\note{Added overview figure with examples showing relationships between Denicek \& co.}
User-centric programming research gave rise to a variety of compelling programming experiences,
including collaborative source code editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming languages and system.

@@

\begin{teaserfigure}
  \vspace{-0.2em}
  \includegraphics[width=1\textwidth,clip,trim=0cm 8.5cm 7.7cm 0cm]{fig/teaser.pdf}
  \vspace{-1.5em}
  \caption{\diff{Denicek is a computational substrate for document-oriented programming
    based on document edit histories~(A). We co-design Denicek with a web-based end-user
    programming environment Denicek (B). Here, Denicek is used to build a Todo list app via
    programming-by-demonstration, by copying a value from input to a new list item.
    We then evaluate the generality of Denicek by using it to build a data science notebook
    system Datnicek (C). Here, Datnicek is used to interactively clean the data table, by
    removing the `` p'' (provisional) marker from numerical columns.
  }}
  \Description{The Denicek diagram in the middle illustrates a document representation as an
   edit history. Two arrows point from the diagram to two screenshots. Webnicek on the left
   contains a Todo list app with a command palette showing add item operation. Datnicek on the
   right contains two cells, one with code to load data and another with an interactive editor
   for data cleaning.}
  \label{fig:teaser}
  \vspace{0.75em}
\end{teaserfigure}
\maketitle

% ==================================================================================================
\section{Introduction}

A computational substrate defines the structures with which programs are constructed, how
the program state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
The choice of a substrate affects what programming experiences can be readily supported.
For example, object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a language laboratory~\cite{steele-1993-lisp}.

@@

\note{Revised to say Denicek is a substrate used to build programming systems.}
\diff{We describe Denicek, a computational substrate that makes it easy to build programming systems
supporting diverse compelling programming experiences \cite{myers-2006-eup}:}
